---
title: "Development actions history"
output: html_document
editor_options: 
  chunk_output_type: console
---

All commands that you use to use when developing packages...

# First time just after creating the project

- Fill the following chunk to create the DESCRIPTION of your package

```{r description, eval=FALSE}
# Describe your package
fusen::fill_description(
  pkg = here::here(),
  fields = list(
    Title = "Spatial Point Pattern Analysis with FDA",
    Version = "0.0.0.9000",
    Description = "Function to analyse TCGA dataset.",
    `Authors@R` = c(
      person("Valerio", "Reffo", email = "valerio.reffo@phd.unipd.it", role = c("aut", "cre"), comment = c(ORCID = "0009-0006-8690-7018"))
    )
  )
, overwrite = TRUE)
# Define License with use_*_license()
usethis::use_gpl3_license()
```

# Start using git

Not required, since we are already on a repository.

```{r, eval=FALSE}
#usethis::use_git()
# Deal with classical files to ignore
#usethis::git_vaccinate()
# Use main for primary branch
#usethis::git_default_branch_rename()
```

# Set extra sources of documentation

```{r, eval=FALSE}
# Install a first time
remotes::install_local()
# README
usethis::use_readme_rmd()
# Code of Conduct
usethis::use_code_of_conduct("contact@fake.com")
# NEWS
usethis::use_news_md()
```

claude suggerisce
```{r, eval=FALSE}
devtools::build_readme()
```


**From now, you will need to "inflate" your package at least once to be able to use the following commands. Let's go to your flat template, and come back here later if/when needed.**

# Inflate flat_read_data.Rmd

Start by inflating `dev/flat_read_data.qmd`:

```{r inflate-read_data, eval=FALSE}
# Run but keep eval=FALSE to avoid infinite loop
# Execute in the console directly
fusen::inflate(flat_file = "dev/flat_read_data.qmd",
               vignette_name = "Reading data",
               overwrite = T)
```

# Create the data for the examples

Needs `read_cells_h5ad()` and `cells2ppp()`, so run it after inflating
`flat_read_data.qmd`. It reads the 5 raw TCGA slides in `data-raw/TCGA/`
(not versioned) and writes:

- `data/tcga_crops.rda` (documented in `flat_read_data.qmd`)
- `inst/extdata/tcga_tiny.h5ad` and `inst/extdata/tcga_tiny_mask.png`

Rerun it only if the raw data or the script change.

```{r example-data, eval=FALSE}
source("data-raw/tcga_examples.R")
```

# Inflate flat_downsampling.qmd

Downsampling functions (`downsample()`, `downsample_uniform()`,
`downsample_KDE()`, `downsample_window()`). The vignette uses `tcga_crops`,
so the example data must exist.

```{r inflate-downsampling, eval=FALSE}
fusen::inflate(flat_file = "dev/flat_downsampling.qmd",
               vignette_name = "Downsampling",
               overwrite = TRUE)
```

# Inflate flat_Lcross_pipeline.qmd

No functions: only the vignette with the pipeline from h5ad files to L-cross
functions. Runs on the example data; the HPC chunks have `eval: false`.

```{r inflate-Lcross_pipeline, eval=FALSE}
fusen::inflate(flat_file = "dev/flat_Lcross_pipeline.qmd",
               vignette_name = "Lcross pipeline",
               overwrite = TRUE)
```

This will put everything in the right place to get started. We then continue by
adding the FDA pipeline (it uses `downsample()`, so inflate it after the
downsampling flat file).

```{r inflate-fda_pipeline, eval=FALSE}
# Run but keep eval=FALSE to avoid infinite loop
# Execute in the console directly
fusen::inflate(flat_file = "dev/flat_fda_pipeline.qmd",
               vignette_name = "FDA pipeline",
               overwrite = TRUE)
```

# Package development tools
## Use once

Only use the commands that you need. You can safely delete those that you never run.

```{r, eval=FALSE}
# Pipe
usethis::use_pipe()

# package-level documentation
usethis::use_package_doc()

# GitHub
# Add your credentials for GitHub
gitcreds::gitcreds_set()
# Send your project to a new GitHub project
usethis::use_github()

# Set Continuous Integration
# _GitHub
# (use_github_action_check_standard() is defunct in usethis >= 2.2)
usethis::use_github_action("check-standard")
usethis::use_github_action("test-coverage")
# pkgdown site on GitHub Pages: creates the gh-pages branch, turns on Pages,
# adds the pkgdown workflow and the URL in DESCRIPTION and _pkgdown.yml
usethis::use_pkgdown_github_pages()
# _GitLab
gitlabr::use_gitlab_ci(type = "check-coverage-pkgdown")

# Add new flat template
fusen::add_flat_template("add")

# Remove renv from the package
usethis::use_build_ignore("renv")
usethis::use_build_ignore("renv.lock")
```

## Use everytime needed

```{r, eval=FALSE}
# Simulate package installation
pkgload::load_all()

# Generate documentation and deal with dependencies
attachment::att_amend_desc()

# Check the package
devtools::check()
```


# Share the package

```{r, eval=FALSE}
# set and try pkgdown documentation website
usethis::use_pkgdown()
pkgdown::build_site()

# build the tar.gz with vignettes to share with others
devtools::build(vignettes = TRUE)
```

